Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-06 article-models-ttmodels.tex
organizing the introduction of section 6

\section{The type theory of models of type theory}
In this section we pursue the idea of what a general model of type theory is by
axiomatizing what you can do with them. We have the following ideas:
\begin{enumerate}
\item There are dependent models and sections thereof. Particular instances
  of sections: extension, weakening and substitution (and something for identity
  function?). And like the original
  extension, weakening and substitution, they're going to be compatible with each
  other. Therefore we state our theory of models as an extension of type theory
  without constructors.
\item Type theory without constructors is a model of itself, the canonical model $\mctx$.
\item for any (family of) model(s) $A$, there is the family model $\mfam{A}$ which
  is a family of models over $A$.
\item The terms of $\mctx$ should be precisely contexts. Terms of $\subst{\Gamma}{\mfam{\mctx}}$
  should be families over $\Gamma$.
\item if a model and a family of models over it are given, there is an extended model.
  If we extend $A$ by $\mfam{A}$, we get the Sierpinski model of $A$.
\item Likewise, models can be weakened and substituted and there are identity
  functions.
\end{enumerate}
So the theory of models is going to be an extension of this theory with this
data. The theory we are about to describe can be seen as an elementary theory
of the category (with families (and terms)) of categories (with families (and terms)).

We would also like to remark explicitly that the valid judgments of the original type
theory become contexts, families or terms, depending on what kind of judgments
it was. Then, valid inferences of the original type theory become valid
judgments here. In some sense, the theory we present here is therefore a second
order theory over the basic theory of types.

@@

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{enumerate}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions. Actually, we want
that every term of this type theory is compatible with those. I.e.~every term
is a functor/morphism of models. Also the sections.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\end{enumerate}
Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.

The easiest set of compatibility rules comes with weakening. The compatibility
rules that deal with substitution are likely going to have to do with the
Yoneda lemma, which we should be able to implement at some point.

The model $\subst{\tfemp{A}}{\mtm{A}}$ is going to be initial in the category of
all models over $A$. That means: whenever $P$ is a family over $\subst{\tfemp{A}}
{\mtm{A}}$ there will be a term of $P$. This means in particular that
$\subst{\tfemp{\emptyf}}{\mtm{\emptyf}}$ is going to be $\mctx$.

>>>>

\section{The type theory of models of type theory}
In this section we pursue the idea of what a general model of type theory is by
axiomatizing what you can do with them. We have the following ideas:
\begin{itemize}
\item There are dependent models and sections thereof. Particular instances
  of sections: extension, weakening and substitution (and something for identity
  function?). And like the original
  extension, weakening and substitution, they're going to be compatible with each
  other. Therefore we state our theory of models as an extension of type theory
  without constructors.
\item Type theory without constructors is a model of itself, the canonical model $\mctx$.
\item for any (family of) model(s) $A$, there is the family model $\mfam{A}$ which
  is a family of models over $A$.
\item The terms of $\mctx$ should be precisely contexts. Terms of $\subst{\Gamma}{\mfam{\mctx}}$
  should be families over $\Gamma$.
\item if a model and a family of models over it are given, there is an extended model.
  If we extend $A$ by $\mfam{A}$, we get the Sierpinski model of $A$.
\item Likewise, models can be weakened and substituted and there are identity
  functions.
\item So the theory of models is going to be an extension of this theory with this
  data. The theory we are about to describe can be seen as an elementary theory
  of the category (with families (and terms)) of categories (with families (and terms)).
\item We would also like to remark explicitly that the valid judgments of the original type
  theory become contexts, families or terms, depending on what kind of judgments
  it was. Then, valid inferences of the original type theory become valid
  judgments here. In some sense, the theory we present here is therefore a second
  order theory over the basic theory of types. 
\end{itemize}

@@

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{itemize}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions. Actually, we want
that every term of this type theory is compatible with those. I.e.~every term
is a functor/morphism of models. Also the sections.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\item Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.
\item The easiest set of compatibility rules comes with weakening. The compatibility
rules that deal with substitution are likely going to have to do with the
Yoneda lemma, which we should be able to implement at some point.
\item The model $\subst{\tfemp{A}}{\mtm{A}}$ is going to be initial in the category of
all models over $A$. That means: whenever $P$ is a family over $\subst{\tfemp{A}}
{\mtm{A}}$ there will be a term of $P$. This means in particular that
$\subst{\tfemp{\emptyf}}{\mtm{\emptyf}}$ is going to be $\mctx$.
\end{itemize}
